Optimal Finite Element Integration

### 2015-07-23 TOMS_Paper/assembly.tex
Overhaul the paper, deleting almost everything (it mostly was shit)

% Page heads
\markboth{F. Luporini et al.}{Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization}

% Title portion
\title{Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization}
\author{Fabio Luporini
\affil{Imperial College London}
David A. Ham
\affil{Imperial College London}
Paul H.J. Kelly
\affil{Imperial College London}}

\begin{abstract}
Abstract goes here
%
%The code is generated from a high level language for PDEs. An
%invocation of the kernel typically does tensor-like calculators over
%relatively small matrices; e.g., triply nested loops operating on
%about a dozen iterations. Because of the DSL context, COFFEE has a lot
%of flexibility on how it organizes/orders the calculations to tailor
%execution to a target architecture.
%
\end{abstract}

@@

\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2014. 
  Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization.}

@@

%Computational cost is a critical limitation in scientific computing,
%especially for finite element simulations. To provide one particular
%example we are particularly concerned about, it has been well
%established that mesh resolution Consequently, our aggressive
%optimization of local assembly, which may even take up to 60% of the
%overall FEM's execution time, directly impacts the performance of
%large-scale scientific simulations running on supercomputers.

\section{Introduction}

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become incredibly popular over the last years. In these frameworks, the weak variational form of a given problem is expressed at high-level by means of a domain-specific language. Such a mathematical specification is suitably manipulated and then passed as input to a form compiler, whose goal is to generate a representation of local assembly operations. These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution. Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more computationally expensive, covering a significant fraction of the overall computation run-time.

@@

Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Finit Element local assEmbly.}, fully integrated with the Firedrake framework~\cite{Luporini}. Besides separating the mathematical domain, captured by the form compiler at an higher level of abstraction, from the optimization process, COFFEE also aims to be platform-agnostic. The code transformations occur on an intermediate representation of the assembly operation, which is ultimately translated into platform-specific code. Domain knowledge is exploited in two ways: for simplifying the implementation of code transformations and to make them extremely effective. Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input. For example, when the input is in the form of an abstract syntax tree produced by the form compiler, specific nodes are decorated so as to drive the optimization process. Although COFFEE has been thought of as a multi-platform optimizing compiler, our performance evaluation so far has been restricted to standard CPU platforms only. We emphasize once more, however, that the transformations applicable by both the Expression Rewriter (Section~\ref{sec:expr-rewriting}) and the Code Specializer (Section~\ref{sec:code-spec}) would work on generic accelerators as well.

In this paper, we build on our previous work~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines based on quadrature representation. We argue that for complex, realistic forms, peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to ensure effective register utilization and SIMD vectorization. The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become steps of code specialization, whereas generalized loop-invariant code motion is performed during the expression rewriting stage. We complement and generalize our previous work with the following contributions.

Expression rewriting aims at reducing the computational intensity of local assembly kernels by rescheduling arithmetic operations based on a set of rewrite rules. Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of operators to expose ``hidden'' loop-invariant sub-expressions and SIMD vectorization opportunities to the code specialization stage. While rewriting an assembly expression, domain knowledge is used in several ways, for example to avoid redundant computation. Secondly, we propose an algorithm that, relying on symbolic execution, restructures the code so as to skip useless iterations over zero-valued entries in vector-valued basis functions arrays, while preserving code vectorizability. These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable using the FEniCS' built-in optimization system, presented in~\cite{quadrature1}.

At code specialization time, transformations are applied to maximize the exploitation of the underlying platform's resources, e.g. SIMD lanes. On top of the work in~\cite{Luporini}, we provide a number of contributions. Firstly, we revisit the padding and data alignment transformation to account for the new code structures produced by expression rewriting. This is of fundamental importance to achieve effective SIMD vectorization. Secondly, we address an open problem in~\cite{Luporini} by providing an algorithm that automatically transforms an element matrix evaluation into a sequence of calls to BLAS' dense matrix-matrix multiply routines. BLAS routines are known to perform far from peak performance when the involved arrays are small, which is almost always the case of low-order finite element methods. However, we will show that in corner, yet meaningful cases, especially in forms characterized by the presence of pre-multiplying functions and relatively high-order function spaces, a BLAS-based implementation can be successful. Finally, we introduce a model-driven, dynamic autotuner that transparently evaluates multiple sets of code transformations to determine the best optimization strategy for a given problem. The main challenge here is to build, for a generic form, a reasonably small search space that comprises most of the effective code variants.

Expression rewriting and code specialization have been implemented in COFFEE, which in turn is integrated with the Firedrake framework. Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on complex hyperelasticity models. We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions. To clearly distinguish the improvement achieved by this work, we will compare, for each test case, four sets of code variants: 1) unoptimized code, i.e. a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e. the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper. Notable performance improvements of 4) over 1), 2) and 3) are reported and discussed.

%The contributions of this paper can be summarized as follows:
%\begin{enumerate}
%\item 
%\item extensive and unprecedented performance evaluation in a wide range of forms of increasing complexity
%\end{enumerate}

%Extensive performance evaluation...

%This paper can be thought of as a continuation of the work presented in~\cite{Luporini}: in particular, we answer the remaining open questions, generalize and formalize an approach to minimize redundant computation using a similar philosophy, we show ~\cite{quadrature1}.

@@

\subsection{Quadrature for Finite Element Local Assembly}
We summarize the basic concepts sustaining the finite element method following the notation adopted in~\cite{quadrature1} and~\cite{Francis}. We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are called bilinear and linear form, respectively. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are computed in the so called assembly phase. Then, in a subsequent phase, the linear system is solved, usually by means of an iterative method, and $\textbf{u}$ is eventually evaluated.

@@

Quadrature schemes are conveniently used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature representation over an element $K$ can be expressed as follows
\begin{equation}
\label{quadrature}
\scriptsize
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

In the next sections, we will often refer to the local element matrix evaluation, such as Equation~\ref{quadrature} for the weighted Lapalce operator, as the \textit{assembly expression} deriving from the weak variational problem.

\subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
\label{sec:summary-opts}
If high performance code needs to be generated, assembly expressions must be optimized with regards to three interrelated aspects: 1) arithmetic intensity, 2) instruction-level parallelism, and 3) data locality. In this paper, we tackle these three points building on our previous work (\cite{Luporini}), in which we presented COFFEE, a mature, platform-independent compiler capable of optimizing local assembly code.

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{pictures/coffee-scheme}
\caption{Outline of COFFEE. The input is either a string of C code or an abstract syntax tree representation. The optimization process goes through two main steps: \textit{expression rewriting} and \textit{code specialization}. The autotuner, incorporated with the Code Specializer, use heuristics to select the set of code variants that, for a given variational form, are likely to reach high performance.}\label{fig:coffee-outline}
\end{center}
\end{figure}

The high-level structure of COFFEE is outlined in Figure~\ref{fig:coffee-outline}. The input is either C code, explicitly provided by the user as a string, or an abstract syntax tree (henceforth AST) representation automatically generated at the higher level of abstraction. We have integrated COFFEE with the Firedrake framework; here, the input is originated by the FEniCS Form Compiler~\cite{FFC-Compiler} in the form of an AST. Decoration of special AST nodes, or equivalently annotation of C code, is extensively used to convey domain knowledge to COFFEE. This is used to introduce more effective, specialized optimizations, as well as to simplify the implementation.

Two conceptually distinct software modules can be individuated: the Expression Rewriter and the Code Specializer. The former targets arithmetic intensity. Its role is to transform the assembly expression and, therefore, the enclosing loop nest, so as to minimize the number of floating point operations that are performed to evaluate the element matrix (or vector). The latter is tailored to optimizing for instruction-level parallelism, particularly SIMD vectorization, and data (register) locality. As shown in the Figure, the Expression Rewriter manipulates and transforms the original AST, which is then provided to the Code Specializer. As described later, we have invested some effort to ensure that the code transformations applied at expression rewriting time would not break any code specialization opportunity. This, for example, would not be the case if the Expression Rewriter coincided with the FEniCS Form Compiler's built-in optimization system; we will clarify this aspect in Section~\ref{sec:zeros}.

To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work. Moreover, in Section~\ref{sec:perf-results}, we will explicitly compare the performance achieved with the transformations presented in this paper to both~\cite{Luporini} and~\cite{quadrature1}.

In our previous work, we have demonstrated the effectiveness of a set of optimizations for finite element quadrature-based integration routines originated through automated code generation. We have also shown how different subsets of optimizations are needed to achieve close-to-peak performance in different variational forms. In particular, we can summarize our study as follows

\begin{itemize}
\item \emph{Generalized Loop-invariant Code Motion}. Compiler's loop-invariant code motion algorithms may not be general enough to optimize assembly expressions, in which different sub-expressions are invariant with respect to more than one loop in the enclosing nest. As reiterated in Section~\ref{sec:expr-rewriting}, we work around this limitation by performing source-level code motion, while also achieving vectorization of invariant code.
\item \emph{Padding and data alignment}. The small size of the loop nest (integration, test, and trial functions loops) require all of the involved arrays to be padded to a multiple of the vector register length so as to maximize the effectiveness of SIMD code. Data alignment can be enforced as a consequence of padding. 
\item \emph{Vector-register Tiling}. Blocking at the level of vector registers, which we perform exploiting the specific memory access pattern of the assembly expressions (i.e. a domain-aware transformation), improves data locality beyond traditional unroll-and-jam optimizations. This is especially true for relatively high polynomial order (i.e. greater than 2) or when pre-multiplying functions are present.
\item \emph{Expression Splitting}. In certain assembly expressions the register pressure is significantly high: when the number of basis functions arrays (or, equivalently, temporaries introduced by loop-invariant code motion) and constants is large, spilling to L1 cache is a consequence for architectures with a relatively low number of logical registers (e.g. 16/32). We exploit sum's associativity to ``split'' the assembly expression into multiple sub-expressions, which are computed individually.
\end{itemize}

In Figure~\ref{fig:coffee-outline}, we show where these transformations are applied in the COFFEE pipeline; also, the novel contributions of this work are highlighted in red. In the next two sections, we describe the new functionalities of Expression Rewriter and Code Specializer, respectively.

%Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.

% Far notare che low-level si, ma sono guidate da high-level knowledge! Questo in risposta alla domanda: ma allora perche' non sfruttare la domain knowledge che hai, che senso ha separare...

% Vantaggio di automatizzare rispetto che scrivere a mano
% which would be practically unfeasible to achieve by hand-written code

\section{Expression Rewriting}
\label{sec:expr-rewriting}
Loop-invariant code motion is the key to reduce the computational intensity of an assembly expression. The Expression Rewriter still relies on this technique, although it makes several steps forward.

Firstly, exploiting operators properties like associativity, distributivity, and commutativity, it manipulates assembly expressions to expose more code hoisting and SIMD vectorization opportunities, as well as to balance the register pressure in the various levels of the loop nest. There are many possibilities of rewriting an expression, so the search space can be quite large. We address this problem by defining a simple yet systematic way of rewriting an expression based on a set of formal rewrite rules, which are meticulously applied by COFFEE.

Secondly, the Expression Rewriter re-structures iteration spaces so as to avoid arithmetic operations over tabulated basis functions columns that are statically known to be zero-valued. Zero-valued columns in tabulated basis functions appear, for example, when taking derivatives on a reference element or when using mixed elements. A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, the issue with that is the use of indirection arrays in the generated code, which break many of the optimizations applicable at code specialization time, including SIMD vectorization. In Section~\ref{sec:zeros}, we present a novel approach based on symbolic execution.

%In COFFEE, the application of the re-writing rules preceds the elimination of operations on zeros.

%It is worth noting how writing these transformations ``by hand'' would be practically unfeasible, especially when the expressions are particularly complex.
%
%
%In this section, we describe our efforts in producing a more effective tool capable of explore how to leverage automated code generation to expose more opportunities to the code hoister. 
%
%Automated code generation, however, Mathematical expressions can be transformed to decrease the number of arithmetic operations required to evaluate the element tensor. A first 
%
%Minimizing the number of floating point operations for evaluating the element tensor of a generic form
%

\subsection{Motivations and Objectives}
\label{sec:expr-rewriter-obj}

\begin{figure}[t]
\tiny
\centering     
\subfigure[Original (simplified) code]{\label{fig:original-code}\lstinputlisting{listings/simple.code}}
~~~~~~~~~~~~~~~~
\subfigure[The code after that generalized loop-invariant code motion has been applied]{\label{fig:invariant-code}\lstinputlisting{listings/invariant.code}}
~\\
~\\
\subfigure[Once terms are factorized, further code hoisting opportunities arise. Here, \texttt{T3} stores the precomputation of an invariant sub-expression exposed by factorization]{\label{fig:factorized-code}\lstinputlisting{listings/factorized.code}}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subfigure[Hoisting and scalar-expansion of integration-dependent terms allow to achieve full SIMD vectorization of the assembly code]{\label{fig:scalarexp-code}\lstinputlisting{listings/vectexpanded.code}}
\caption{Applying a sequence of transformations to an excerpt from a Burgers' local assembly kernel, suitably simplified to improve readability.}
\end{figure}

Consider the local element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a Burgers problem. The assembly expression, produced by the FEniCS Form Compiler, has been deliberately simplified, and code details have been omitted for brevity and readability. In practice, assembly expressions can be much more complex, for example depending on the differential operators employed in the variational form; however, this example is representative enough for highlighting patterns that are common in a large class of problems.

A first glimpse of the code suggests that the sub-expression \texttt{a*f0*A[i][j]+b*f1*B[i][j]} is invariant with respect to the innermost (trial functions) loop \texttt{k}, so it can be hoisted at the level of the outer loop \texttt{j} to avoid redundant computation. This is indeed a standard compiler transformation, supported by any available compilers, so, in principle, there should be no need to transform the source code explicitly. With a closer look we notice that the sub-expression \texttt{d*D[i][k]+e*E[i][k]} is also invariant, although, this time, with respect to the outer (test functions) loop \texttt{j}. In~\cite{Luporini}, we have showed that available compilers limit the search for code motion opportunities to the innermost loop of a given loop nest. Moreover, the lack of cost models to ascertain both the optimal place where to hoist an expression and whether or not vectorizing it at the price of extra temporary memory is a fundamental limiting factor. The \textit{generalized loop-invariant code motion} technique that we implemented in COFFEE, leading to the code in Figure~\ref{fig:invariant-code}, addressed these limitations, which are critical in the case of non-trivial assembly expressions, resulting in run-time improvements up to 3$\times$.

In this paper, we target particularly complex variational forms for which a more aggressive transformation is required. We generalize the problem of finding and hoisting invariant sub-expressions by formulating the following critical question: \textit{how an assembly expression should be rewritten so as to minimize both floating point operations and data movement}? To minimize floating point operations, in terms of instructions executed, it is important to reduce the absolute number of arithmetic operations to evaluate the element matrix as well as to achieve full vectorization of the assembly loop nest.

We start considering the case of assembly expressions that, after generalized loop-invariant code motion has been applied, still  ``hide'' opportunities for code hoisting. By examining again the code in Figure~\ref{fig:invariant-code}, we notice that the basis function array \texttt{A} iterating along the \texttt{[i,j]} loops appears twice in the expression. By expanding the products in which \texttt{A} is accessed and by applying sum commutativity, terms can be factorized. This has two effects: firstly, it reduces the number of arithmetic operations performed; secondly, and most importantly, it exposes a new sub-expression \texttt{A[i][k]/c+T2[k]*f} invariant with respect to loop \texttt{j}. Consequently, hoisting can be performed, resulting in the code in Figure~\ref{fig:factorized-code}. In general, exposing factorization opportunities requires traversing the whole expression tree, and then expanding and moving terms. It also needs heuristics to select a factorization strategy: there may be different opportunities of reorganizing sub-expressions, and, in our case, the best is the one that maximizes the invariant code eventually disclosed. We will discuss this aspect formally in Section~\ref{fig:rewrite-rules}.

As a second observation, we note that integration-dependent expressions are inherently executed as scalar code. For example, the \texttt{f0*a} and \texttt{f1*b} products in Figure~\ref{fig:factorized-code} depend on the loop along quadrature points; these operations are performed in a non-vectorized way at every \texttt{i} iteration. This is not a big issue in our running example, in which the scalar computation represents a small fraction of the total, but it becomes a concrete problem in complicated forms, like those at the heart of hyperelasticity (which we evaluate in Section~\ref{sec:perf-results}). In such forms, the amount of computation independent of both test and trial functions loops is so large that it has a significant impact on the run-time, despite being executed only $O(I)$ times (with $I$ number of quadrature points). We have therefore implemented an algorithm to move and scalar-expand integration-dependent expressions, which leads to codes as in Figure~\ref{fig:scalarexp-code}.

The third motivation for expression rewriting concerns the register pressure in the innermost loop. Once the code has been optimized for arithmetic intensity, it is important to think about how the transformations impacted register allocation. Assume the local assembly kernel is executed on a state-of-the-art CPU architecture having 16 logical registers, e.g. an Intel Haswell. Each value appearing in the expression is loaded and kept in a register as long as possible. In Figure~\ref{fig:scalarexp-code}, for instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every \texttt{i} iteration. This implies that at every iteration of the \texttt{[j,k]} loop nest, 12$\%$ of the available registers are spent just to store values independent of test and trial functions loops. In more complicated expressions, the percentage of registers destined to store such constant terms can be even higher. Registers are, however, a precious resource, especially when evaluating compute-intensive expressions. The smaller is the number of available free registers, the worse is the instruction-level parallelism achieved: for example, a shortage of registers can increase the pressure on the L1 cache (i.e. it can worsen data locality), or it may prevent the effective application of standard transformations, e.g. loop unrolling. We aim at relieving this problem by suitably expanding terms and introducing, where necessary, additional temporary values. We illustrate this in the following example.

%An analogous analysis applies to processors with larger numbers of registers, since using loop unroll or loop unroll-and-jam to expose more instruction-level parallelism would increase the requirements on registers.

\begin{figure}[t]
\tiny
\centering
\subfigure[Original (simplified) code]{\label{fig:toexpand-code}\lstinputlisting{listings/toexpand.code}}
~\\
\subfigure[The code after expansion of \texttt{det*W[i]}]{\label{fig:expanded-1-code}\lstinputlisting{listings/expanded-1.code}}
~~
\subfigure[The code after expansion of \texttt{g}. Note the need to introduce a new temporary array]{\label{fig:expanded-2-code}\lstinputlisting{listings/expanded-2.code}}
\caption{Expansion of terms to improve register pressure in a local assembly kernel}\label{fig:expanded-code}
\end{figure}

Consider a variant of the Burgers local assembly kernel, shown in Figure~\ref{fig:toexpand-code}. This is again a representative, simplified example. We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of the invariant sub-expression stored in \texttt{T1}, resulting in code as in Figure~\ref{fig:expanded-1-code}. Freeing the register destined to the constant \texttt{g} is less straightforward: we cannot absorb it in \texttt{T1} as we did with \texttt{det*W[i]} since \texttt{T1} is also accessed in the \texttt{T1[j]*A[i][k]} sub-expression. The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}. Generalizing, this is a problem of data dependencies: to solve it, we use a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}. The dependency graph is initially empty, and is updated every time a new temporary is created by either loop-invariant code motion or expansion of terms. The dependency graph is then queried to understand when expansion can be performed without resorting to new temporary values. This aspect is formalized in the next section.

\subsection{Rewrite Rules}
\label{sec:rewrite-rules}
In general, assembly expressions produced by automated code generation can be much more complex than those we have used as examples, with dozens of terms involved (basis function arrays, derivatives, coefficients, ...) and hundreds of (nested) arithmetic operations. Our goal is to establish a portable, unified, platform-independent, and systematic way of reducing the computational strength of an expression exploiting the intuitions described in the previous section. This \textit{expression rewriting} should be simple; definitely, it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems, such as COFFEE. In other words, we look for an algorithm capable of transforming a plain assembly expression by applying 1) generalized loop-invariant code motion, 2) non-trivial factorization and re-association of sub-expressions, and 3) expansion of terms; after that, it should perform 4) scalar-expansion of integration-dependent terms to achieve full vectorization of the assembly code.

We centre such an algorithm around a set of rewrite rules. These rules drive the transformation of an expression, prescribe where invariant sub-expressions will be moved (i.e. at what level in the loop nest), and track the propagation of data dependencies. When applying a rule, the state of the loop nest must be updated to reflect, for example, the use of a new temporary and new data dependencies. We define the state of a loop nest as $L = (\sigma, G)$, where $G = (V, E)$ represents the dependency graph, while $\sigma : Inv \rightarrow S$ maps invariant sub-expressions to identifiers of temporary arrays. The notation $\sigma_i$ refers to invariants hoisted at the level of loop $i$. We also introduce the \textit{conditional hoister} operator $[]$ on $\sigma$ such that
\begin{gather*}
\sigma[\sfrac{v}{x}] = 
\begin{cases}
\sigma(x) \text{~~~~~~~~if $x \in Inv$; $v$ is ignored}\\
v \text{~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $\sigma$(x) = v}\\
\end{cases}
\end{gather*}
That is, if the invariant expression \texttt{x} has already been hoisted, $[]$ returns the temporary identifier hosting its value; otherwise, \texttt{x} is hoisted and a new temporary $v$ is created. There is a special case when $v = \perp$, used for \textit{conditional deletion} of entries in $\sigma$. Specifically
\begin{gather*}
\sigma[\sfrac{\perp}{x}] = 
\begin{cases}
\sigma(x) \text{~~~~~~~~if $x \in Inv$; $\sigma = \sigma \setminus (x, \sigma(x))$}\\
v \text{~~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $v \notin S$}\\
\end{cases}
\end{gather*}
In other words, the invariant sub-expression \texttt{x} is removed and the temporary identifier that was hosting its value is returned if \texttt{x} had been previously hoisted; otherwise, a fresh identifier $v$ is returned. This is useful to express updates of hoisted invariant sub-expressions when expanding terms.

In the following, a generic (sub-)expression is represented with Roman letters \texttt{a, b, ...}; constant terms are considered a special case, so Greek letters \texttt{$\alpha$, $\beta$, ...} are used instead. The iteration vector $i = [i_0, i_1, ...]$ is the ordered sequence of the indices of the loops enclosing an (sub-)expression. We will refer to $i_0$ as the outermost enclosing loop. The notation $a_i$, therefore, indicates that the expression $a$ assumes distinct values while iterating along the loops in $i$; its outermost loop is silently assumed to be $i_0$.

>>>>

% Page heads
\markboth{F. Luporini et al.}{Optimal Finite Element Integration}

% Title portion
\title{Optimal Finite Element Integration}
\author{Fabio Luporini
\affil{Imperial College London}
David A. Ham
\affil{Imperial College London}
Paul H.J. Kelly
\affil{Imperial College London}}

\begin{abstract}
...
\end{abstract}

@@

\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. Optimal Finite Element Integration.}

@@

\section{Introduction and Motivations}

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become increasingly popular over the last years. In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language. The mathematical specification is manipulated and then passed to a form compiler, which generates a representation of local assembly operations. These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution. Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more computationally expensive, covering a significant fraction of the overall computation run-time.

@@

Our research has resulted in ....
... we build on our previous work~\cite{Luporini} ... and present a ... We argue that for complex, realistic forms, peak performance can be achieved only by  ... 
... also low-level optimisation ...

This is all implemented in COFFEE, which in turn is integrated with the Firedrake framework. We provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on complex (hyperelasticity) models. We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions. To clearly distinguish the improvement achieved by this work, we will compare, for each test case, X sets of code variants: 1) unoptimized code, i.e. a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e. the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; ....

@@

We summarize the basic concepts sustaining the finite element method following the notation adopted in~\cite{quadrature1} and~\cite{Francis}. We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are called bilinear and linear form, respectively. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are computed in the so called assembly phase. Then, in a subsequent phase, the linear system is solved, usually by means of an iterative method, and $\textbf{u}$ is eventually evaluated.

@@

\subsection{Quadrature Mode}
Quadrature schemes are conveniently used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature representation over an element $K$ can be expressed as follows
\begin{equation}
\label{quadrature}
\scriptsize
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

\subsection{Tensor Mode}
...

\section{Optimality of Multilinear Loop Nests}
\label{sec:optimal-impl}

We need to define: 
\begin{itemize}
\item multilinear loop nests, static loop nests; (note: the loop over the elements could also be considered static, if mesh didn't move...)
\item concept of reuse in a loop;
\item prove that reuse can always be eliminated from the innermost loop of multilinear loop nests. 
\item definition of optimality
\item insights that optimisation of loops outside of the multilinear nest can be tackled greedily and heuristically
\end{itemize}

Need to show examples.

\section{Automated Code Generation of Optimal Multilinear Loop Nests}

\begin{itemize}
\item Deficiencies of previous approaches
\item  Building block operations in coffee: 1) eliminating reuse, 2)  insights on the various tree algorithms employed
\item cost model for deciding what to do
\item constrained increase in the working set size
\item heuristic optimisation of constant and quadrature dependent expressions
\end{itemize}

\section{Low-level Optimization}
\label{sec:code-spec}
...

\subsection{Avoiding Iteration on Zero-valued Blocks by Symbolic Execution}
\label{sec:zeros}
%TODO : say that the reasons to have zeros is to keep the design of the form compiler simple, which just "puts zeros" and "outline the assembly expression".

Operations over blocks of zero-valued entries in tabulated (derivatives of) basis functions can be skipped. Zero-valued columns arise, for example, when taking derivatives on a reference element, or when employing vector function spaces. In~\cite{quadrature1}, a technique to avoid iteration over zero-valued columns, based on the use of indirection arrays (e.g. \texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}), was described and tested in \texttt{FFC}. We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}. Our strategy differs by avoiding indirection arrays in the generated code.
